Comparing probabilistic predictive models applied to football

### 2015-07-08 template.tex
Correcoes.

	The famous saying of John von Neumann: {\it ``With four parameters I can fit an elephant, and with five I can make him wiggle his trunk''} is a perhaps just a humorous version of Ockham's razor:
	{\it``Entities are not to be multiplied without necessity''}, but illustrates very well a principle that is not often followed by researchers, especially when building predictive models.
	This is true also in the literature of probabilistic sports predictions, particularly regarding models devoted to football or soccer\footnote{Henceforth, football.} predictions.

@@

	The predictions of the two benchmark models were published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2). 
	We know that\footnote{One of the authors is responsible for one of these models and the other was described in a dissertation. See \cite{arruda2000}.} both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength for each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
	Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models\footnote{To be precise we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.}.

@@

	Formally, let $X$ be a random variable
	taking values in $\mathcal{X}=\{1,2,3\}$ indicating
	the outcome of the match, with 1 standing for home win, 2 for draw and 3 for away win.  Moreover, let $P=(P_1,P_2,P_3)$ denote one's probabilistic prediction
	about $X$, i.e., $P$ lies in the 2-simplex set $\Delta_2=\{(p_1,p_2,p_3):p_1+p_2+p_3=1, \ p_1,p_2,p_3\geq0\}$ (see Figure
	\ref{fig:simplex}). 	
	A scoring rule is a function 
	that assigns a real number (score) $S(x,P)$ to each $x \in \mathcal{X}$
	and $P \in \Delta_2$ 
	such that 
	for any given $x$ in $\mathcal{X}$, the score  $S(x,P)$ is minimized when $P$ is 
	$(1,0,0)$, $(0,1,0)$ or $(0,0,1)$ depending if $x$ is 1, 2 or 3, respectively.
	The score $S(x,P)$ can be thought as
	a penalty to be paid when one assigns the 
	probabilistic prediction $P$ and outcome
	$x$ occurs. Also, the ``best''\footnote{A scoring rule can also be defined to be such that a \emph{large} value of the score indicates good forecasts.} possible score (i.e., the smallest score value) is achieved when the probabilistic prediction for the outcome of the game is perfect.
	%Most of the scoring rules also exhibit an arbitrary normalizing feature such as $S(x,P_X)=0$ for each $x$.

@@

	Figure~\ref{fig::scores} displays the boxplots of the scores and proportion of errors of the predictive models considered.
	%It also shows an horizontal line with the score of the trivial prediction.
	According to all scoring rules, all methods have similar performance, and they are more accurate than the trivial prediction, displayed in the plots
	as an horizontal line\footnote{None of the 95\% confidence intervals for the mean score contain the score given by the trivial prediction.}. The mean scores and their standard errors are displayed
	in Table \ref{tab::brier}.

@@

	Figure \ref{fig::entropy} presents the boxplots of the entropy of each prediction for the predictive models considered.\footnote{Recall
		that the entropy of a prediction $(p_1,p_2,p_3)$ is given by $- \sum_{i=1}^3 p_i \log{p_i}$}
	Since the entropy values are smaller than 1.09, the plots show that the predictions
	are typically more informative than the trivial prediction. Nevertheless, all methods yielded similar entropies in average, that is,
	none of them seem to provide more informative probabilities.

@@

	%De Finetti was a radical subjectivist and pragmatist about probability and did not believe that it even made sense to talk about credences as being ``accurate'' or ``inaccurate''.
	%Like Ramsey and Savage, he offered pragmatic arguments for probabilism.
	%As a result, de Finetti did not speak of ``measures of inaccuracy'' or ``measures of distance from the truth'': he used the terms ``scoring rule'' and ``loss function''.

	%De Finetti used what is called the Brier score, which is just the (squared) Euclidean distance ? i.e., the sum of the squared differences between credences and (the 0/1 numerical correlates of) truth values.

	%In the statistical literature, football prediction models can be divided in two main approaches: (i) modeling the number of goals \citep{Maher82, Dixon97, Karlis2003} and (ii) modeling the final outcome (win, draw or lose) of the matches \citep{Forrest2000, Koning2000, Brillinger2008, Brillinger2009}. For a discussion of these two approaches see \cite{Goddard2005}.
	Although our benchmark models were here defined as complex, we can find much more elaborate models in the literature, hindering the estimation and prediction processes.
	Among them, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to
	change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}.

	%Although the benchmark models were here considered complex models, we can find in the literature much more elaborate models, hindering the estimation and prediction process. 
	%Among those, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}.

	According to standard scoring rules and other criteria here adopted to qualify the predictions, our findings show that the predictive power of the three simple models was quite similar to the benchmark models.
	%We compared the probabilistic predictions of three simple models with the ones provided by two more complex models for the second round of the 2014 Brazilian football championships of first and second divisions. 
	This work thus indicates that while there has been a boom in the literature with the proposal of several fancy and involved models, especially in the era of big data, simple models can many times do
	as well as their complex counterparts.

	This is often unnoticed by users of such methods, which leads to the overuse of models that are
	hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity}\footnote{In allusion to Anderson's {\it The
			Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''.}, that is not, of course, always observed. 
	In this spirit, since we have explored only the prediction of football matches, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.

>>>>

	The famous saying of John von Neumann: {\it ``With four parameters I can fit an elephant, and with five I can make him wiggle his trunk''} is a perhaps just a humorous version of Ockham's razor:
	{\it``Entities are not to be multiplied without necessity''}, but illustrates very well a principle that is not often followed by researchers, especially when building predictive models.
	This is true also in the literature of probabilistic sports predictions, particularly regarding models devoted to football or soccer (henceforth football) predictions.

@@

	The predictions of the two benchmark models were published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2). 
	Since one of the authors is responsible for one of these models and the other was described in a dissertation \cite{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength for each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
	Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models. To be precise we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.

@@

	Formally, let $X$ be a random variable
	taking values in $\mathcal{X}=\{1,2,3\}$ indicating
	the outcome of the match, with 1 standing for home win, 2 for draw and 3 for away win.  Moreover, let $P=(P_1,P_2,P_3)$ denote one's probabilistic prediction
	about $X$, i.e., $P$ lies in the 2-simplex set $\Delta_2=\{(p_1,p_2,p_3):p_1+p_2+p_3=1, \ p_1,p_2,p_3\geq0\}$ (see Figure
	\ref{fig:simplex}). 	
	A scoring rule is a function 
	that assigns a real number (score) $S(x,P)$ to each $x \in \mathcal{X}$
	and $P \in \Delta_2$ 
	such that 
	for any given $x$ in $\mathcal{X}$, the score  $S(x,P)$ is minimized when $P$ is 
	$(1,0,0)$, $(0,1,0)$ or $(0,0,1)$ depending if $x$ is 1, 2 or 3, respectively.
	The score $S(x,P)$ can be thought as
	a penalty to be paid when one assigns the 
	probabilistic prediction $P$ and outcome
	$x$ occurs. Also, the ``best'' possible score (i.e., the smallest score value) is achieved when the probabilistic prediction for the outcome of the game is perfect. A scoring rule can also be defined to be such that a \emph{large} value of the score indicates good forecasts.
	%Most of the scoring rules also exhibit an arbitrary normalizing feature such as $S(x,P_X)=0$ for each $x$.

@@

	Figure~\ref{fig::scores} displays the boxplots of the scores and proportion of errors of the predictive models considered.
	%It also shows an horizontal line with the score of the trivial prediction.
	According to all scoring rules, all methods have similar performance, and they are more accurate than the trivial prediction, displayed in the plots
	as an horizontal line. None of the 95\% confidence intervals for the mean score contain the score given by the trivial prediction. The mean scores and their standard errors are displayed
	in Table \ref{tab::brier}.

@@

	Figure \ref{fig::entropy} presents the boxplots of the entropy of each prediction for the predictive models considered. Recall that the entropy of a prediction $(p_1,p_2,p_3)$ is given by $- \sum_{i=1}^3 p_i \log{p_i}$.
	Since the entropy values are smaller than 1.09, the plots show that the predictions
	are typically more informative than the trivial prediction. Nevertheless, all methods yielded similar entropies in average, that is,
	none of them seem to provide more informative probabilities.

@@

	Although our benchmark models were here defined as complex, we can find much more elaborate models in the literature, hindering the estimation and prediction processes.
	Among them, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}.
	However, models build to predict the number of goals must predict a more specific outcome and this usually requires more elaborate modeling.

	According to standard scoring rules and other criteria here adopted to qualify the predictions, our findings show that the predictive power of the three simple models was quite similar to the benchmark models.
	This work thus indicates that while there has been a boom in the literature with the proposal of several fancy and involved models, especially in the era of big data, simple models can many times do as well as their complex counterparts.

	This is often unnoticed by users of such methods, which leads to the overuse of models that are
	hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity}, in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. However , we emphasize that this phenomenon is not always observed. 
	In this spirit, since we have explored only the prediction of football matches, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.
